---
title: "Appendix"
output:
  pdf_document:
    highlight: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```





All files used in this document can be found at: https://github.com/bodkan/admixr-SI.




# Estimating the proportions of Neandertal ancestry



Our goal is to estimate Neandertal ancestry in some present-day European
individuals using the following $f_4$-ratio statistic:

$$\alpha = \frac{f_4(\textrm{Altai Neandertal, Chimpanzee; European X, African})}{f_4(\textrm{Altai Neandertal, Chimpanzee; Vindija Neandertal, African})}.$$

We will demonstrate how to accomplish this using:

a) a traditional ADMIXTOOLS workflow;

b) a new workflow with the _admixr_ R package.


\pagebreak


## a) A traditional ADMIXTOOLS workflow

To perform this analysis using a traditional ADMIXTOOLS workflow we have to do
the following:

1. Create a `pop` file (`f4ratio.pop`).

This file specifies the order of populations in the $f_4$ statistic calculation
above, and will contain the following (one $f_4$-ratio setup per row, as
required by ADMIXTOOLS):

```{bash, echo = FALSE}
cat f4ratio.pop
```

2. Create a `par` file (`f4ratio_all.par`).

This is the main configuration file and contains the paths to a trio of EIGENSTRAT
components and a path to the `pop` file created in the previous step:

```{bash, echo = FALSE}
cat f4ratio_all.par
```

3. Run the `qpF4ratio` command-line program and capture its output:


```{bash}
qpF4ratio -p f4ratio_all.par > f4ratio_all.log
```

4. Inspect the output file (`f4ratio_all.log`) and extract the relevant values.

To extract the values of interest we could either copy-paste them manually into
Excel, or extract them using a combination of grep/awk/sed and save them into a
text file for further analysis (we will omit these steps in this Appendix).

\pagebreak

## b) A new workflow using the _admixr_ package

We can accomplish the same task by running the following R code, which will
perform all the low-level configuration and parsing work automatically,
returning a simple R dataframe object.

```{r}
library(admixr)

results <- f4ratio(
  X = c("French", "Sardinian", "Czech"),
  A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp",
  data = eigenstrat("snps")
)

results
```



\pagebreak


# Estimate Neandertal ancestry proportions on a subset of the data.

Let's investiate a variation of the analysis above, this time estimating
Neandertal ancestry proportions in a subset of the genome (in a set of regions
specified in a BED file `regions.bed`).

## a) A traditional ADMIXTOOLS workflow

All ADMIXTOOLS commands accept a parameter `badsnpname: <snp file>` in a `par`
file, which directs ADMIXTOOLS to ignore SNPs at specified coordinates.

In a traditional ADMIXTOOLS workflow we would have to create this file using
some form of shell scripting. For example, we could do the following:

1. Convert the EIGENSTRAT snp file into a BED format:

```{bash}
awk -v OFS='\t' '{print $2, $4 - 1, $4, $0}' snps.snp > snps.bed
```

2. Intersect the newly generated BED file with coordinates of regions in
`reigons.bed`, keeping those sites that fall outside of those regions and
saving them in a `snp` format again:


```{bash}
bedtools intersect -v -a snps.bed -b regions.bed | cut -f 4- > excluded_sites.snp
```

3. Create a new `par` file called `f4ratio_subset.par`, setting the `badsnpname`
parameter to `excluded_sites.snp`:

```{bash, echo = FALSE}
cat f4ratio_subset.par
```

4. Run `qpF4ratio` with the new `par` file and capture its output:

```{bash}
qpF4ratio -p f4ratio_subset.par > f4ratio_subset.log
```

5. Extract the relevant values from the output file (`f4ratio_subset.log`).




\pagebreak




## b) A new workflow using the _admixr_ package



Using the _admixr_ package, we can use the function `filter_bed()` to do the
filtering work without the need for cumbersome and error-prone shell scripting:

```{r}
library(admixr)

snps <- eigenstrat("snps") # trio of EIGENSTRAT files is: snps.{ind,snp,geno}
subset <- filter_bed(snps, "regions.bed") # create a new EIGENSTRAT object

results <- f4ratio(
  X = c("French", "Sardinian", "Czech"),
  A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp",
  data = subset
)

results
```



Furthermore, because all _admixr_ functions are inspired by the "tidyverse"
philosophy, we can utilize the `%>%` pipe operator from the `magrittr` package
and run the whole analysis in one step:

```{r}
library(admixr)
library(magrittr)

results <-
  eigenstrat("snps") %>% # trio of EIGENSTRAT files is: snps.{ind,snp,geno}
  filter_bed("regions.bed") %>% # create a new EIGENSTRAT object and pass it further
  f4ratio(
    X = c("French", "Sardinian", "Czech"),
    A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp"
  )

results
```





\pagebreak



# A more complex example

The "pipe-friendly" nature of _admixr_ makes it possible to compose even more
complex pipelines, such as the following:

```{r}
library(admixr)
library(magrittr)

results <-
  eigenstrat("snps") %>% # trio of EIGENSTRAT files is: snps.{ind,snp,geno}
  relabel(Europeans = c("French", "Sardinian", "Czech")) %>%
  filter_bed("regions.bed") %>%
  transversions_only() %>%
  f4ratio(X = "Europeans", A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp")

results
```

Accomplishing the same taks using pure shell scripting would require
a significant amount of work.