---
title: "Supplementary Information"
output:
  word_document:
    highlight: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```





All files used in this document and its source in R Markdown can be found at:
https://github.com/bodkan/admixr-SI.

A complete description of the functionality of the _admixr_ package can be found
in the tutorial vignette at: https://bodkan.net/admixr/articles/tutorial.html




# Example analysis: estimating the proportions of Neandertal ancestry



Our goal is to estimate the proportions of Neandertal ancestry in a set of
present-day European individuals (French, Sardinian and Czech) using the
following $f_4$-ratio statistic:

$$\alpha = \frac{f_4(\textrm{Altai Neandertal, Chimpanzee; European X, African})}{f_4(\textrm{Altai Neandertal, Chimpanzee; Vindija Neandertal, African})}.$$

We will demonstrate how to perform this analysis using:

a) a traditional ADMIXTOOLS workflow;

b) a new workflow with the _admixr_ R package.



## a) A traditional ADMIXTOOLS workflow

To perform this task using a traditional ADMIXTOOLS workflow, we have to do
the following:

1. Create a `pop` file (`f4ratio.pop`).

This file specifies the order of populations in the $f_4$ statistic calculation
above, and will contain the following (one $f_4$-ratio setup per row, as
required by ADMIXTOOLS):

```{bash, echo = FALSE}
cat f4ratio.pop
```

2. Create a `par` file (`f4ratio_all.par`).

This is the main configuration file and contains the paths to a trio of EIGENSTRAT
components and a path to the `pop` file created in the previous step.

```{bash, echo = FALSE}
cat f4ratio_all.par
```

3. Run the `qpF4ratio` command-line program and capture its output:

```{bash}
qpF4ratio -p f4ratio_all.par > f4ratio_all.log
```

4. Inspect the output file (`f4ratio_all.log`) and extract the relevant values.

To extract the values of interest (on lines beginning with "result:"), we could
either copy-paste them manually into Excel, or extract them using a combination
of grep/awk/sed and save them into a text file for further analysis (we will
omit these steps here).

```{bash, echo = FALSE}
cat f4ratio_all.log
```

## b) A new workflow using the _admixr_ package

Running the following R code will perform all the low-level configuration and
parsing work (steps 1-4 above) automatically, returning a simple R data frame:

```{r}
library(admixr)

result <- f4ratio(
  X = c("French", "Sardinian", "Czech"),
  A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp",
  data = eigenstrat("data/snps")
)

result
```


# Estimate Neandertal ancestry proportions on a subset of the data.

Let's investiate a variation of the analysis above, this time estimating
Neandertal ancestry proportions in a subset of the genome (in a set of regions
specified in a BED file `regions.bed`).

## a) A traditional ADMIXTOOLS workflow

All ADMIXTOOLS commands accept a parameter `badsnpname`, which can be specified
in a `par` file and instructs ADMIXTOOLS to ignore SNPs at specified
coordinates.

In a traditional ADMIXTOOLS workflow, we would have to create this file using
some form of shell scripting. For example, we could do the following:

1. Convert the EIGENSTRAT snp file into a BED format:

```{bash}
awk -v OFS='\t' '{print $2, $4 - 1, $4, $0}' data/snps.snp > snps.bed
```

2. Intersect the newly generated BED file with coordinates of regions in
`reigons.bed`. We want to keep only those sites that fall outside of those
regions and saving them in a `snp` format again.

```{bash}
bedtools intersect -v -a snps.bed -b regions.bed \
    | cut -f 4- \
    > excluded_sites.snp
```

3. Create a new `par` file (`f4ratio_subset.par`), with a new parameter
`badsnpname`:

```{bash, echo = FALSE}
cat f4ratio_subset.par
```

4. Run `qpF4ratio` with the new `par` file and capture its output:

```{bash}
qpF4ratio -p f4ratio_subset.par > f4ratio_subset.log
```

5. Extract the relevant values from the output file (`f4ratio_subset.log`):

```{bash, echo = FALSE}
cat f4ratio_subset.log
```



## b) A new workflow using the _admixr_ package


Using the _admixr_ package, we can use the function `filter_bed()` to do
all filtering automatically:

```{r}
library(admixr)

snps <- eigenstrat("data/snps")
subset <- filter_bed(snps, "regions.bed") # create a new EIGENSTRAT object

result <- f4ratio(
  X = c("French", "Sardinian", "Czech"),
  A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp",
  data = subset
)

result
```


Furthermore, because all _admixr_ functions are inspired by the "tidyverse
philosophy", we can utilize the `%>%` pipe operator from the `magrittr` package
and run the whole analysis in one step:

```{r}
library(admixr)
library(magrittr)

result <-
  eigenstrat("data/snps") %>%
  filter_bed("regions.bed") %>%
  f4ratio(
    X = c("French", "Sardinian", "Czech"),
    A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp"
  )

result
```




# A more complex example

The "pipe-friendly" nature of _admixr_ makes it possible to compose even more
complex pipelines, such as the following:

```{r}
library(admixr)
library(magrittr)

result <-
  eigenstrat("data/snps") %>%
  relabel(Europeans = c("French", "Sardinian", "Czech")) %>%
  filter_bed("regions.bed") %>%
  transversions_only() %>%
  f4ratio(
    X = "Europeans",
    A = "Altai", B = "Vindija", C = "Yoruba", O = "Chimp"
  )

result
```

Accomplishing the same taks using pure shell scripting would require
a significant amount of work.